---
title: "stringsAsFactors = HELLNO"
author: "Jenny Bryan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{stringsAsFactors = HELLNO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Factors are a very useful type of variable in R, but they can also drive you nuts. Especially the "stealth factor" that you think of as character.

Can we soften some of their sharp edges?

Binding two factors via `fbind()`:

```{r}
library(foofactors)
a <- factor(c("character", "hits", "your", "eyeballs"))
b <- factor(c("but", "integer", "where it", "counts"))
```

Simply catenating two factors leads to a result that most don't expect.

```{r}
c(a, b)
```

The `fbind()` function glues two factors together and returns factor.

```{r}
fbind(a, b)
```

Often we want a table of frequencies for the levels of a factor. The base `table()` function returns an object of class `table`, which can be inconvenient for downstream work.  Processing with `as.data.frame()` can be helpful but it's a bit clunky.

```{r}
set.seed(1234)
x <- factor(sample(letters[1:5], size = 100, replace = TRUE))
table(x)
as.data.frame(table(x))
```

The `freq_out()` function returns a frequency table as a well-named `tbl_df`:

```{r}
freq_out(x)
```

By default, factor levels are ordered by alphabet. However, sometimes we want to order the levels in a different way, for example when we want to plot data in the order of increasing or descreasing values. Let's have a look at the median of `Sepal.Width` for each `Species` in the `iris` data:

```{r, warning = FALSE}
library(ggplot2)

ggplot(iris, aes(Species, Sepal.Width)) + 
  geom_boxplot() +
  labs(x = "")
```

We can change the order of levels with the `reorder` function from the `stats` package, which orders levels of a factor based on the values in another variable.

```{r}
Species_reordered <- reorder(iris$Species, iris$Sepal.Width, median)

ggplot(iris, aes(Species_reordered, Sepal.Width)) + 
  geom_boxplot() +
  labs(x = "")
```

However, `reorder` always orders in ascending order. What if we want it ordered descendingly? The new `foofactors` function `reorder_desc` can help us with that:


```{r}
Species_reordered_desc <- reorder_desc(iris$Species, iris$Sepal.Width, median)
ggplot(iris, aes(Species_reordered_desc, Sepal.Width)) + 
  geom_boxplot() +
  labs(x = "")
```

Note that in the above examples `median` was specified as the summary statistic of `Sepal.Width` on which the order of `Species` will be based. If no function is specified, the default used is `mean`.
